---
title: "3D brain organoid model reconstructions from 2D bright field images using R"
author: "Alik Huseynov"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    theme: lumen
vignette: >
  %\VignetteIndexEntry{make3dOrganoid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Some setups for chunks
```{r setups, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, results='hide')
```

#### Dependencies

install the relevant libs from R CRAN or from their github repo 
```{r load_libs}
# load the libraries
library(Morpho)
library(rgl)
library(parallel)
library(Arothron)
library(ctrlR)
library(deformetrics)
library(Rvcg)
library(abind)
library(mesheR)
```

set some R options
```{r set_options}
options(max.print=5.5E5) # can visualize huge array
options(scipen = 500) # standard notation vs scientific one!
options(digits = 6) # digits after comma
#rm(list = ls()) # useful to remove everything from the current environment
mc.cores<-parallel::detectCores() # detect how many core the computer has
```

#### Import segmented organoid outlines (image processing and segmentation was done in Fiji).
define the segmented outlines, check workflow_manual.pdf - step1 Fiji and - step2 R. Make sure you are in in the right working directory. Eg "/Users/YOUR_USER_NAME/Desktop/make3dOrganoid_rg/output_fiji/"
```{r import_2d_outlines}
# my working directory
setwd("/Users/Work/Desktop/make3dOrganoid_rg/output_fiji/")
# fig1
group_f1<-paste0("fig1_",c("w4_mutant","w4_control","w6_mutant","w6_control","w8_mutant","w8_control"))
ind_f1<-seq(6)

# fig2
group_f2<-paste0("fig2_",c("d4_mutant","d4_control","w4_mutant","w4_control","w8_mutant","w8_control"))
ind_f2<-seq(6)
# list files and remove .txt extension
file_lists<-c(list.files("fig1f_Li_etal17_input.jpg/",pattern = ".txt",full.names = F),
                 list.files("fig2b_Li_etal17_input.jpg/",pattern = ".txt",full.names = F))
file_lists<-sapply(seq(length(file_lists)), function(i) gsub(".txt","",file_lists[i])) # make numeric
file_list<-c(list.files("fig1f_Li_etal17_input.jpg/",pattern = ".txt",full.names = T),
                list.files("fig2b_Li_etal17_input.jpg/",pattern = ".txt",full.names = T))
# note which organoid belogns to which stage and group, check the fig2 in /ref_paper4data
outline_list<-lapply(seq(length(file_list)), function(i) as.matrix.data.frame(read.table(file_list[i])))
## subsample and make equidistant
str(outline_list)
# for dimnames
c(group_f1,group_f2)

# minumum outline points
min_outline_size<-min(sapply(seq(length(outline_list)), function(i) dim(outline_list[[i]])[1]))
# make equidistant
set.seed(3332)
outline_list_eqdist<-list()
for (i in seq(length(outline_list))) {
  outline_list_eqdist[[i]]<-equidistantCurve(outline_list[[i]][sort(sample(dim(outline_list[[i]])[1],size = min_outline_size,replace = F)),],
                                             iterations = 2,open = F,smoothit = 1,increment = 1)}
# remove #268 point as it is the same as #1
for (i in seq(length(outline_list))){outline_list_eqdist[[i]]<-outline_list_eqdist[[i]][-268,]}
```

#### Generate equidistant 3d curves/outlines
```{r plot_3d_outline}
# add a 3rd "z" dimention (eg 0)
outline_list_eqdist_3dcurve<-list()
for (i in seq(length(outline_list))) {outline_list_eqdist_3dcurve[[i]]<-cbind(outline_list_eqdist[[i]],0)}
#have a look at 3d curve
outline_list_eqdist_3dcurve[[12]] %>% spheres3d(radius = 1)
```

#### Do GLS/Procrustes registration (linear transformations) on 3d curves
```{r gls1}
# make an array
outline_list_eqdist_3dcurve_arr<-list2array(outline_list_eqdist_3dcurve)
# add dimnames
dimnames(outline_list_eqdist_3dcurve_arr)[[2]]<-c("x","y","z")
dimnames(outline_list_eqdist_3dcurve_arr)[[3]]<-c(group_f1,group_f2)
# you can do partial or full procrustes registration
proc_outline3d_partial<-procSym(outline_list_eqdist_3dcurve_arr,scale = F,CSinit = F) # no scaling
proc_outline3d_full<-procSym(outline_list_eqdist_3dcurve_arr) # full
dim(proc_outline3d_partial$orpdata) # 267 points, 3 dimensions xyz, and 12 organoids/specimens
# look at points number
open3d(); text3d(proc_outline3d_partial$orpdata[,,1],texts = as.character(seq(dim(proc_outline3d_partial$orpdata[,,1])[1])))
text3d(proc_outline3d_partial$orpdata[,,2],texts = as.character(seq(dim(proc_outline3d_partial$orpdata[,,2])[1])),col="red")
#
#replicate(length(rgl.dev.list()),rgl.close())
```


#### Make 3d shapes out of the 3d curves
```{r make_3d_shapes}
# you would need to test several steps for organoid outlines from your own images
###
# step 1 replicate and scale the outlines at some custom steps
# steps for elevation and demotion or lowering (cast down vs cast up)
steps2updown<-rev(c(.98,.90,.75,.55,.30)) # custom best/default steps
open3d(); invisible(lapply(seq(length(steps2updown)), function(i) spheres3d(proc_outline3d_partial$orpdata[,,1]*steps2updown[i],radius = .3,col="black")))
#rgl.clear()
# make a function to generate the 3d shape(s), the output would be a matrix.
# some form of extrusion with scaling each outline using steps2updown vector
# the amount of distance for elevation and demotion (amount_dist4elevation_numeric):
# will be the average dist. of pairwise point distances (distance matrix or euclidean distance matrix) / by a value you define
makeShape3d<-function(curve_outline3d_matrix,steps2updown_vector,amount_dist4elevation_numeric){
  final_matrix_3dshape<-NULL
  # scaling
  scaled_listUp<-list()
  for (i in seq(length(steps2updown_vector))){
    scaled_listUp[[i]]<-curve_outline3d_matrix*steps2updown_vector[[i]]
  }
  # a list for lowering/cast down
  scaled_listDown<-scaled_listUp
  # eg. use half of the average euclidean distace as maximum for elevation
  FM<-function(M){as.matrix(dist(M))} # from helper functions https://gist.github.com/benmarwick/7651452 Morphometrics with R Claude J.2008
  distMatEucld<-FM(curve_outline3d_matrix)
  #mean(colMeans(distMatEucld)) # average max. value for elevation
  #mean(colMeans(distMatEucld))/amount_dist4elevation_numeric
  # steps for cast down
  elevateSteps<-seq(mean(colMeans(distMatEucld))/amount_dist4elevation_numeric,
                    c(mean(colMeans(distMatEucld))/amount_dist4elevation_numeric)/length(steps2updown_vector),
                    length.out = length(steps2updown_vector))
  # cast up
  for (i in seq(length(steps2updown_vector))){
    scaled_listUp[[i]][,3]<-elevateSteps[i]
  }
  # lowering of the curves or demotion/cast down
  for (i in seq(length(steps2updown_vector))){
    scaled_listDown[[i]][,3]<--elevateSteps[i]
  }
  # final 3d shape as matrix
  final_matrix_3dshape<-rbind(deformetrics::list2matrix(scaled_listUp),
                              curve_outline3d_matrix,
                              deformetrics::list2matrix(scaled_listDown))
  return(final_matrix_3dshape)
}
#
test_3dshape<-makeShape3d(proc_outline3d_partial$orpdata[,,1],steps2updown,amount_dist4elevation_numeric = 1.4)
open3d(); spheres3d(test_3dshape,radius = .5)
# NOTE: higher values for amount_dist4elevation_numeric will result in elongated 3d model, lower values for flat 3d model.
# use 1.4 (turns out to be the best) or 1.3 for the amount of mean distance for elevation and lowering.
proc_outline3d_partial_shape3d<-
  lapply(seq(dim(proc_outline3d_partial$orpdata)[3]), 
         function(i) makeShape3d(proc_outline3d_partial$orpdata[,,i],steps2updown,amount_dist4elevation_numeric = 1.4))
# check
# names
names(proc_outline3d_partial$size)
#
open3d(); spheres3d(proc_outline3d_partial_shape3d[[1]],radius = 1) # fig1 week4 mutant
open3d(); spheres3d(proc_outline3d_partial_shape3d[[2]],radius = 1,col="red") # fig1 week4 control
#
#replicate(length(rgl.dev.list()),rgl.close())
```

#### Do GLS/Procrustes registration (linear transformations) on 3d shapes
```{r gls2}
# make a final array
outline3d_arr<-list2array(proc_outline3d_partial_shape3d)
dim(outline3d_arr)
# add dimnames
dimnames(outline3d_arr)[[2]]<-dimnames(proc_outline3d_partial$orpdata)[[2]]
dimnames(outline3d_arr)[[3]]<-dimnames(proc_outline3d_partial$orpdata)[[3]]
# partial procrustes registration
proc_outline3d_arr<-procSym(outline3d_arr,scale = F,CSinit = F) # no scaling
dim(proc_outline3d_arr$orpdata) # 2937 points, 3 dimensions xyz, and 12 organoids/specimens
# optionally - look at points number
#open3d(); text3d(proc_outline3d_arr$orpdata[,,1],texts = as.character(seq(dim(proc_outline3d_arr$orpdata[,,1])[1]))) # fig1 week4 mutant
#text3d(proc_outline3d_arr$orpdata[,,2],texts = as.character(seq(dim(proc_outline3d_arr$orpdata[,,2])[1])),col="red") # fig1 week4 control
#replicate(length(rgl.dev.list()),rgl.close())
```

#### Triangulate the organoid 3d shape to generate template 3d mesh model
here one spherical ogranoid 3d shape will be used to generate template 3d mesh model
```{r mesh_template3d, message=T}
# check 3d shape of specimen #12
library(geometry)
dimnames(outline3d_arr)[[3]][12]
proc_outline3d_arr$orpdata[,,12] %>% spheres3d(radius = 1) # pick more spherical or round organoid as a template
# triangulation
inputriang<-t(surf.tri(proc_outline3d_arr$orpdata[,,12], 
                       delaunayn(proc_outline3d_arr$orpdata[,,12],options = "QJ")))
rgl.triangles(proc_outline3d_arr$orpdata[,,12][inputriang,1], 
              proc_outline3d_arr$orpdata[,,12][inputriang,2], 
              proc_outline3d_arr$orpdata[,,12][inputriang,3], col = "ivory", alpha=0.5)
#
rgl.close()
# save the model as stl
dimnames(outline3d_arr)[[3]][12]
stl_fig2_w8_control<-paste0("/Users/Work/Desktop/make3dOrganoid_rg/",dimnames(outline3d_arr)[[3]][12],".stl")
open3d()
rgl.triangles(proc_outline3d_arr$orpdata[,,12][inputriang,1], 
              proc_outline3d_arr$orpdata[,,12][inputriang,2], 
              proc_outline3d_arr$orpdata[,,12][inputriang,3], col = "ivory", alpha=0.5)
writeSTL(stl_fig2_w8_control)
rgl.close()
# load it and remesh
setwd("/Users/Work/Desktop/make3dOrganoid_rg/")
proc_fig1_w4_control_mesh_or<-vcgUpdateNormals(vcgIsolated(vcgImport("fig2_w8_control.stl")))
dim(proc_fig1_w4_control_mesh_or$vb)
proc_fig1_w4_control_mesh<-
  rmUnrefVertex(vcgClean(vcgIsolated(vcgUniformRemesh(vcgImport("fig2_w8_control.stl"),
                                                      voxelSize = 15,multiSample = T)),sel = c(1,2,3,4,1)))
dim(proc_fig1_w4_control_mesh$vb)
shade3d(vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200),lit=T,col="gray70",specular="black")
shade3d(vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200),lit=T,col="white",front="lines",back="lines")
# exclude 2 curves eg -c(1:267,1603:1869)
transferPoints(proc_outline3d_arr$orpdata[-c(1:267,1603:1869),,12], 
               vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200),
               vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200)) %>% spheres3d(radius = 1)
# similar method of transfering the points onto a mesh
#vert2points(vcgClostKD(proc_outline3d_arr$orpdata[,,12],vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200))) %>% spheres3d(radius = 1)
#
replicate(length(rgl.dev.list()),rgl.close())
```

Smoothly transfrom the template organoid 3d mesh into all organoids in the sample (TPS transform)
```{r make3d_organoids}
# mesh list of all the specimens
# some surface smoothing as well
# 3dshape of the template mesh, specimen12
# and exclude 2 curves eg -c(1:267,1603:1869). This helps to overcome the "plateau effect" on top-bottom of the 3d model
proc_outline3d_arr$orpdata[-c(1:267,1603:1869),,8] %>% spheres3d(radius = 1)
temp_meshShape_sp12<-transferPoints(proc_outline3d_arr$orpdata[-c(1:267,1603:1869),,12],
               vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200),
               vcgSmooth(rmUnrefVertex(proc_fig1_w4_control_mesh),lambda = .8,iteration = 200))
proc_organoids_meshes<-mclapply(seq(length(proc_outline3d_arr$size)), function(i) 
  tps3d(vcgSmooth(proc_fig1_w4_control_mesh,lambda = .8, iteration = 200),temp_meshShape_sp12,proc_outline3d_arr$orpdata[-c(1:267,1603:1869),,i]),mc.cores =  mc.cores-2)
#
# list of all organoids
dimnames(proc_outline3d_arr$orpdata)[[3]]
# have a look at the first specimen
# "fig1_w4_mutant"
shade3d(proc_organoids_meshes[[1]],lit=T,col="gray70",specular="black")
shade3d(proc_organoids_meshes[[1]],lit=T,col="white",front="lines",back="lines")
#
replicate(length(rgl.dev.list()),rgl.close())

# visualize 3d organoids
open3d(windowRect=c(80,80,800,800))
shade3d(proc_organoids_meshes[[1]],lit=T,col="gray70",specular="black"); shade3d(proc_organoids_meshes[[1]],lit=T,col="white",front="lines",back="lines")
# close all the rgl windows
replicate(length(rgl.dev.list()),rgl.close())

# optionally, to plot each organoid
#dimnames(proc_outline3d_arr$orpdata)[[3]] # names
#for (i in seq(length(proc_organoids_meshes))) { open3d()
#  shade3d(proc_organoids_meshes[[i]],lit=T,col="gray70",specular="black");shade3d(proc_organoids_meshes[[i]],lit=T,col="white",front="lines",back="lines")}
#
#replicate(length(rgl.dev.list()),rgl.close())
#============================================================================
```

#### Do final GLS/Procrustes registration (linear transformations) on all the 3d mesh verices
```{r gls3}
# how many vertices 
dim(proc_organoids_meshes[[1]]$vb) # number of points
# extract mesh vertices
proc_organoids_meshes_lms<-lapply(seq(length(proc_organoids_meshes)), function(i) vert2points(proc_organoids_meshes[[i]]))
proc_organoids_meshes_lms<-list2array(proc_organoids_meshes_lms)
# add dimnames
dimnames(proc_organoids_meshes_lms)[[2]]<-dimnames(proc_outline3d_arr$orpdata)[[2]]
dimnames(proc_organoids_meshes_lms)[[3]]<-dimnames(proc_outline3d_arr$orpdata)[[3]]
# do full procrustes (PCA included)
proc_organoids_meshes_lms_proc<-procSym(proc_organoids_meshes_lms)

# save data if needed
#save.image("~/Desktop/make3dOrganoid_rg/make3dOrganoid.RData")
#============================================================================
```

#### Calculate 3d organoid surface complexitry using Dirichlet normal energy
this is done using the package molaR (originally developed to study teeth morphology).
```{r cal_surface_complexity}
# surface complexity quantification
library(molaR)
# get the shape metric, i.e Dirichlet normal energy of a surface
# ie the bending of the surface, high DNE indicates that a surface has ridges - ridged surface
derichetDNE_meshes<-mclapply(seq(length(proc_organoids_meshes)), function(i) 
  DNE(proc_organoids_meshes[[i]],BoundaryDiscard = "None"),mc.cores = mc.cores-2)

# names for each organoids
dimnames(proc_organoids_meshes_lms_proc$orpdata)[[3]]
# visualize each organoid 3d model
DNE3d(derichetDNE_meshes[[12]],edgeMask = F,outlierMask = F,leftOffset = .2)
rgl.clear()

# optionally, get the relief index of the surface, plus 2d and 3d area
?molaR::RFI
#relief2d3dareas_meshes<-mclapply(seq(length(proc_hu_nea_organoid_meshes)), function(i) RFI(proc_hu_nea_organoid_meshes[[i]],alpha = 0.5),mc.cores = 3)
# visualize
#RFI3d(relief2d3dareas_meshes[[1]])
# rgl.close()
##
```

#### Calculate 3d organoid size
here we can calculate varous size statistics, such as centroid size mesh volume.
```{r cal_organoid_size}
# centroid size of the mesh vertices was already calculated during GLS using Morpho::procSym()
proc_organoids_meshes_lms_proc$size
# calculate mesh volumes
vol_proc_organoids_meshes<-lapply(seq(length(proc_organoids_meshes)), function(i) vol.mesh(proc_organoids_meshes[[i]])) %>% list2matrix()
dimnames(vol_proc_organoids_meshes)[[1]]<-names(proc_organoids_meshes_lms_proc$size)
log10(vol_proc_organoids_meshes)
```

#### Plot DNE, to see how organoid surface complexity chages during development
```{r plot_DNE, warning=F}
# DNE statistic can be found here:
derichetDNE_meshes[[12]]$Surface_DNE # eg for organoid #12
surf_complex<-sapply(seq(length(proc_organoids_meshes)), function(i) derichetDNE_meshes[[i]]$Surface_DNE)
dimnames(proc_organoids_meshes_lms_proc$orpdata)[[3]][7:12]
# make df for plot
df4plot<-data.frame(cbind(round(surf_complex[7:12],1), 
                          round(log10(vol_proc_organoids_meshes),1), 
                          dimnames(proc_organoids_meshes_lms_proc$orpdata)[[3]][7:12]),
                    substring(dimnames(proc_organoids_meshes_lms_proc$orpdata)[[3]][7:12],9,15),
                    substring(dimnames(proc_organoids_meshes_lms_proc$orpdata)[[3]][7:12],6,7))
dimnames(df4plot)[[2]]<-c("DNE", "log10_mesh_volumes", "names","genotypes","devo_stages")
df4plot
library(ggplot2)
ggDNE<-ggplot(data = df4plot, aes(x = devo_stages, y = DNE, color = genotypes)) + 
  geom_line(linetype = "dashed", alpha = 0.8, aes(group = genotypes)) + 
  geom_point(shape=19, alpha = 0.8, aes(col=genotypes, size=log10_mesh_volumes), show.legend = T) + 
  #scale_colour_manual(values = c("green3", "red")) + 
  #scale_alpha(range = c(0.1, 0.2)) + 
  ggplot2::theme(
            # Remove panel grid lines
            panel.grid.major = ggplot2::element_blank(),
            panel.grid.minor = ggplot2::element_blank(),
            # Remove panel background
            panel.background = ggplot2::element_blank()) + 
  labs(subtitle="Control vs Mutant", 
       y="DNE", x="Developmental stages", 
       title="Changes in organoid surface complexity (DNE) and size (mesh volumes)")
plot(ggDNE)
```
